\subsection{Time-varying magnetic field}
    A concuctive mesh with resistance $R$ will, in the presence of an external magnetic field \B, 
    have an induced current $I_i$ such that
    \begin{equation}
        \emf_\text{i} = RI_\text{i} = -\frac{\partial \Phi(\B)}{\partial t},
    \end{equation}
    where $\emf_\text{i}$ is the induced electromotive force, and $\Phi(\B)$ is the flux of \B{} through the mesh. 
    Any change in the flux will result in an induced current, 
    be it by rotating or stretching the mesh (changing the projected area),
    moving the mesh or source, or change the field intensity (if the source has such properties).

    Since an $\emf$ must be accompanied by an electric field, 
    there will be an induced electric field $\E_\text{i}$ such that
    \begin{equation}
        \oint_\mathcal{L}\E_\text{i}\cdot d\vec{l} = -\frac{\partial \Phi(\B)}{\partial t},
    \end{equation}
    valid for \textit{any} line $\mathcal{L}$, regardless of path or medium. 
    Therefore, this field is present in vacuum and requires no conductors.

    Note that $\E_\text{i}$ is NOT conservative in general. The differential form of the equation is
    \begin{equation}
        \nabla \times \E_\text{i} = -\frac{\partial \B}{\partial t}.
    \end{equation}
    Note that this also holds in general, since $\nabla\times\E = 0$ for an electrostatic field \E.

\subsection{Time-varying electric field} \label{Time-varying electric field}
    A changing electrical field induces a \textit{displacement current} such that
    \begin{equation}
        \nabla \cdot \j_s = -\nabla\left(\e\frac{\partial\E}{\partial t}\right),
    \end{equation}
    where $\j_s$ is the \textit{displacement current density}. This is not a real current in the sense that charges are moving, 
    but it behaves as a current density.

    $\j_{\text{tot}} = \j_{\text{conduction}} + \j_s$ is always solenoidal. 

    Note that 
    \begin{equation*}
        I_s = \Phi(\j_s) = \iint_S \e\frac{\partial\E}{\partial t}\cdot d\vec{S} = \e\frac{\partial}{\partial t}\Phi_S(\E)
    \end{equation*}

    Inserting this into Ampere's law, we get 
    \begin{align}
        \oint_{\partial S}\B\cdot d\vec{l} &= \mu_0 I_{\text{enclosed}} + \mu_0\e\frac{\partial}{\partial t}\Phi_S\left(\E\right)\\
        \nabla\times\B &= \mu_0\j_{\text{concuctive}} + \mu_0\e\frac{\partial\E}{\partial t},
    \end{align}
    where we often omit the subscript on $\j$, and $\partial S$ is the perimiter of the surface $S$.

\subsection{Examples of magnetically induced current}
    \subsubsection*{Rod on a rail}
        Two conductive parallel rails, seperated by a distance $b$ and connected with a resistor $R$, stretch infinitley long, 
        and is immersed in a static and uniform \B-field (i.e. constant in time and space). 
        A conductive rod with resistance $r$ is placed between the rods at a distance $x_0$ from the resistor, 
        connecting the rods and forming a mesh. The rod is then given a velocity $v$ along the rails, away from the resistor.
        \B{} points parallel to the normal vector of the mesh. We assume zero friction. 
        The flux of the magnetic field through the mesh is given by
        \begin{equation}
            \Phi(\B)(t) = Bb(vt+x_0) \rightarrow -\frac{\partial \Phi(\B)}{\partial t} = -Bbv.
        \end{equation}
        Therefore, the induced current in the mesh is
        \begin{equation}
            I_\text{i} = -\frac{Bbv}{R+r}.
        \end{equation}
        Since all the values are positive, the minus sign dictates the direction of the current. Applying the right hand rule,
        the thumb should therefore point antiparallel to \B. 
        (More rigorously, the flux was calculated with the assumption of a surface normal parallel to \B. 
        The minus sign therefore flips the direction of current.)

        This induced current creates an opposing force (opposite direction of $v$): 
        \begin{equation}
            \vec{F}_\text{i} = -\frac{B^2b^2}{R+r}\vec{v}.
        \end{equation}
        It is clear that, if no external force is applied, the rod will stop moving.

    \subsubsection*{Rotating mesh}
        A rectangular mesh (resistance $R$) with area $S$ rotates along an axis along its surface with angular velocity $\omega$. 
        It is immersed in a static, uniform \B-field (i.e. constant in time and space).
        We get 
        \begin{align}
            \Phi(\B)(t) &= BS\cos (\omega t), \\
            \emf_\text{i} &= \omega S B \sin (\omega t).
        \end{align}
        The instantaneous and average dissapated power is 
        \begin{align}
            P_\text{i} &= \frac{\omega^2 B^2 S^2}{R}\sin^2(\omega t),\\
            \left<P_\text{i}\right> &= \frac{1}{T}\int_0^T P_\text{i}dt = \frac{\omega^2 B^2 S^2}{2R},
        \end{align}
        where $T = \frac{2\pi}{\omega}$ is the period of a turn.
        Since power is dissapated, a torque is required to keep the mesh spinning: 
        \begin{equation}
            \vec{\tau}_\text{ext} = \vec{m}\times\B = \frac{\omega B^2 S^2}{R}\sin^2(\omega t).
        \end{equation}
    
    \subsubsection*{Self-induction}
        A circuit with current induces a magnetic field. 
        If the current varies, then the flux of the induced magnetic field through the circuit also varies. 
        We get
        \begin{equation}
            \frac{\partial \Phi(\B)}{\partial t} = L_{1,1}\frac{\partial I}{\partial t} = -\emf_{\text{i}},
        \end{equation}
        where $L_{1,1}$ is the \textit{coefficient of self-inductance}. (see \ref{Magnetic flux} for mutual inductance)
\subsection{Transient circuits}
    \subsubsection{RL-circuits}
        The simplest RL-circuits consist of a source $\emf$, a resistor $R$, and a coil (solenoid) $L$.
        Here, we add a switch as well.

        Applying Kirchoff's first law, we arrive at the differential equation
        \begin{equation}
            \emf - RI = L_{1,1}\frac{\partial I}{\partial t},
        \end{equation}
        solved by
        \begin{equation}
            \emf - RI = Ae^{\frac{-R}{L}t}
        \end{equation}
        where $A$ is an integration constant and is determined by the initial conditions.

        With initial conditions $t=0,\, I=0$, we get 
        \begin{equation}
            I(t) = \frac{\emf}{R}\left(1-e^{\frac{-L}{R}t}\right),
        \end{equation}
        i.e. closing the switch at $t=0$ results in an inverse exponential increase in current 
        with an upper bound of $\frac{\emf}{R}$, i.e. conductor-like behaviour for $L$.

        Consider initial conditions $t=0\, I = \frac{\emf}{R}$, i.e. we open the switch at $t=0$.
        We now consider the switch as a resistor with $R' >> R$, since a spark will still allow current for a while.
        \begin{equation}
            I(t) = \frac{\emf}{R'}\left(1-e^{\frac{-L}{R+R'}t}\right) + \frac{\emf}{R} e^{\frac{-L}{R+R'}t},
        \end{equation}
        which is closely approximated (since $R+R' \approx R'$ and $R'^{-1} \approx 0$) by
        \begin{equation}
            I(t) \approx \frac{\emf}{R}e^{\frac{-L}{R'}t}.
        \end{equation}
        The resulting behaviour of the coil is then to keep the current flowing for a while, but the decay is exponential.

    \subsubsection{RLC-circuits}
        A system with a resistor with resistance $R$, a coil with coefficient of self induction $L$, and a capacitor with capacitance $C$, 
        all in series, is described by the following differential equation: 
        \begin{equation} \label{Current in RLC}
            \frac{\partial^2}{\partial t^2}I + \frac{R}{L}\frac{\partial}{\partial t}I + \frac{1}{LC}I = 0.
        \end{equation}
        We define the \textit{damping coefficient} $\gamma$ and the \textit{eigen frequency} $\omega_0$ as follows:
        \begin{align*}
            \gamma &= \frac{R}{2L}\\
            \omega_0 &= \sqrt{\frac{1}{LC}},
        \end{align*}
        yielding in the boring exponential solutions for $\gamma^2 \geq {\omega_0}^2$:
        \begin{equation}
            I(t) = e^{-\gamma t}\left(Ae^{\sqrt{\gamma^2 - {\omega_0}^2}t}+Be^{-\sqrt{\gamma^2 - {\omega_0}^2}t}\right),
        \end{equation}
        where $A$ and $B$ are integration coefficients. The system is more interesting for $\gamma^2 \leq {\omega_0}^2$,
        where the complex exponential yields oscilating behaviour:
        \begin{equation}
            I(t) = e^{-\gamma t}\cdot A\sin\left(\omega t + \phi\right),
        \end{equation}
        where $A$ and $\phi$ are integration constants, and $\omega = \sqrt{\gamma^2 - {\omega_0}^2} > 0$.

        With $R=0$, there will be no energy dissapated over the resistor, and we get 
        \begin{equation*}
            E = E_c + E_L = \frac{1}{2}C{V_c(t)}^2 + \frac{1}{2}L{I(t)}^2,
        \end{equation*}
        where the current $I$ and the potential difference between the capacitor plates $V_c$ are given by 
        \begin{align*}
            I(t) &= A\sin\left(\frac{1}{\sqrt{LC}}t + \phi\right) \\
            V_c (t) &= \sqrt{\frac{L}{C}} \, A\cos\left(\frac{1}{\sqrt{LC}}t + \phi\right)
        \end{align*}

    \subsubsection{Permanent oscilations in RLC-circuits}

        Adding an AC source $\emf(t) = \emf_0\cos\left(\omega t + \phi\right)$ to our RLC-circuit, where all components are in parallel, 
        we get equation \eqref{Current in RLC} equal to $\frac{-\omega\emf_0}{L}\sin\left(\omega t + \phi\right)$.
        
        We wave our hands, and get the following as a solution: 
        \begin{equation}
            I(t) = I_0 \cos\left(\omega t\right),
        \end{equation}
        where 
        \begin{align*}
            \phi &= \text{arctan} \left(\frac{\omega L - \left(\omega C\right)^{-1}}{R}\right)\\
            I_0 &= \frac{\emf_0}{\sqrt{R^2 + \left(\omega L - \left(\omega C\right)^{-1}\right)^2}},
        \end{align*}
        where $\phi$ is the phase of the AC source.

        Plotting $I_0(\omega)$ (usually) yields a bell-curve. We define $\Delta\omega$ as the width of the curve at $I_0 = \frac{\emf_0}{\sqrt{2}R}$: 
        \begin{align*}
            \omega_1 &= \frac{-R}{2L} + \sqrt{\frac{R^2}{4L^2} + {\omega_0}^2}\\
            \omega_2 &= \frac{R}{2L} + \sqrt{\frac{R^2}{4L^2} + {\omega_0}^2}\\
            \Delta\omega &= \frac{R}{L},
        \end{align*}
        where $\omega_0 = \frac{1}{\sqrt{LC}}$.

        Note that $\omega_1\omega_2={\omega_0}^2$, and $|\omega_1 - \omega_0| \neq |\omega_2 - \omega_0|$, i.e. the curve is asymmetric.

        $\Delta\omega$ is used for the \textit{quality factor}:
        \begin{equation}
            Q = \frac{\omega_0}{\Delta\omega}.
        \end{equation}

        This is useful for sensors where the measurement is produced by a change in capacitance; 
        if $Q$ is large then we get a narrow range of frequencies with substantial current, and a changed capacitance 
        (resulting in a changed $\omega_0$) is easy to notice. 

        A small $Q$, on the other hand, will result in a wide band of "accepted" frequencies, which can be useful in other circumstances.
    
        Which size ranges are considered "small" and "large" was not discussed during lectures. 

\subsection{Electromagnetic waves}
    \subsubsection{Magnetic vector potential}
        Since \B{} is solenoidal, it can be written as the curl of another vector field. 
        
        Let $\vec{A}$ be a vector field, such that
        \begin{equation}
            \nabla \times \vec{A} = \B.
        \end{equation}
        $\vec{A}$ is not a physical field, but it is a helpful mathematical tool. 

        Manipulating Maxwell's equations, we arrive at 
        \begin{equation}
            \E = -\nabla V - \frac{\partial \vec{A}}{\partial t}.
        \end{equation}
        
        $\vec{A}$ can be changed by adding the gradient of any arbitrary scalar function, 
        however, to keep the measurable fields \B{} and \E{} unchanged, one also needs to change $V$ correspondingly:
        \begin{align*}
            \vec{A}' &= \vec{A} + \nabla S \\
            V' &= V - \frac{\partial S}{\partial t}
        \end{align*}
        where $S$ is a scalar function with both a spatial- and a time-parameter.
        Applying this mathematical feature is called a \textit{Gauge transform}.
    \subsubsection*{Arriving at the wave equation}
        Inserting $V$ and $\vec{A}$ into Maxwell's equations, we get the following: 
        \begin{align*}
            \nabla^2V - \e\mu_0\frac{\partial^2V}{\partial t^2} &= \frac{-\rho}{\e}\\
            \nabla^2\vec{A} - \e\mu_0\frac{\partial^2\vec{A}}{\partial t^2} &= \mu_0\j,
        \end{align*}
        which is clearly the wave equation if there are no sources present ($\rho = \j = 0)$, 
        where the velocity of propagation is $c$ (since $\e\mu_0 = c^{-2}$).

        We use the shorthand notation of the \textit{d'Alembertian}: 
        \begin{equation*}
            \dalambert = \nabla^2 - \frac{1}{c^2}\frac{\partial^2}{\partial t^2}
        \end{equation*}

        and arrive at
        \begin{center}
            \begin{tabular}{ c|c }
                $\dalambert V = 0$&$ \dalambert\,\E = 0$\\
                $\dalambert\,\vec{A} = 0 $&$\dalambert\,\B = 0$
            \end{tabular}
        \end{center}
    \subsubsection{Solution to the wave equation}
        The solution  can be written as
        \begin{equation*}
            f(\vec{r}, t) = A\sin\left(\vec{k}\cdot\vec{r} - \omega t\right),
        \end{equation*}
        where $A$ is the amplitude, $|\vec{k}| = \frac{2\pi}{\lambda}$ is the wave "number", pointing in the direction of travel, 
        and $\omega = 2\pi\nu$ is the angular frequency. $\lambda$ is the wavelength and $\nu$ is the frequency, such that
        $\lambda\nu = c$.

        Both \E, \B{} and $\vec{k}$ are orthogonal, i.e. electromagnetic waves are transverse. Also note that 
        \begin{equation}
            E = cB
        \end{equation}

    \subsubsection{Energy of an electromagnetic wave}
        Summing the electric and magnetic components yields the following energy density: 
        \begin{equation}
            u_{\text{em}} = \e E^2 = \frac{1}{\mu_0}\, B^2.
        \end{equation}
    
    \subsubsection{Poynting vector}
        The Poynting vector describes the power of an electromagnetic wave per area:
        \begin{equation}
            \vec{S} = \frac{1}{\mu_0}\,\E\times\B.
        \end{equation}
        $\vec{S}$ is measured in \si{\joule\per\second\metre\squared}.
        The average length of $\vec{S}$ can be found by multiple methods:
        \begin{equation}
            \left<|\vec{S}|\right> = \frac{1}{T}\int_T Sdt = \frac{1}{2}\e c\E^2 = \frac{1}{2\mu_0}c\B^2
        \end{equation}
    
\subsection{Polarisation of electromagnetic waves}
    There are two types of polarisation discussed in the lectures: linear and eliptical.
    By convention, the polarisation refers to the electric field, 
    but the magnetic field, being orthogonal, will behave similarly.
    
    Most non-electrical light-sources produce \textit{unpolarised light}, which, contrary to what the name suggests, 
    contains both linearly and eliptically polarised light in somewhat equal proportions. 

    \subsubsection*{Linearly polarised electromagnetic waves}
        Both fields of the electromagnetic waves only oscilate in one direction,
        i.e. if the wave propagates in the $z$-direction, 
        then the phase difference between the $x$- and $y$-component of each field is zero.
        Mathematically, the wave can be written as
        \begin{equation}
            \E(z, t) = E_{0x}\sin\left(kz - \omega t + \phi\right)\hat{x} +
            E_{0y}\sin\left(kz - \omega t + \phi\right)\hat{y},
        \end{equation}
        where $E_{0x,y}$ are the decomposed amplitude of the wave, 
        $k$ is the wave vector (one-dimensional in this case), 
        $\omega$ is the angular frequency, 
        and $\phi$ is an arbitrary initial phaseshift which is usually set to 0.

    \subsubsection*{Eliptically polarised electromagnetic waves}
        Eliptically polarised waves propagating in the $z$-direction can be mathematically described by
        \begin{equation}
            \E(z, t) = E_{0x}\sin\left(kz - \omega t + \phi_x \right)\hat{x} +
            E_{0y}\sin\left(kz - \omega t + \phi_y \right)\hat{y},
        \end{equation}
        where $E_{0x,y}$ are the decomposed amplitude of the wave, 
        $k$ is the wave vector (one-dimensional in this case), 
        $\omega$ is the angular frequency, 
        and $\phi$ is an arbitrary initial phaseshift. 
        In this course, only $\phi_x = \frac{\pi}{2}$ and $\phi_y = 0$ is discussed.
        
        The phase difference of the two components yield a helical behavior, 
        as compared to the linear behaviour of the linearly polarised counterpart.

        Note that, if $E_{0x} = E_{0y}$, (and $\phi_x = \phi_y + \frac{\pi}{2}$, which is true for this course)
        then the elipse becomes a circle, and the wave is \textit{circularly polarised}.

        Applying the right hand rule, 
        one can determine if the polarised wave is a right-hand or left-hand circularly polarised.

\subsection{The electromagnetic spectrum}
    The following table shows some basic information for the different classifications of electromagnetic waves.
    \begin{longtable}[c]{@{}c|c@{}} 
        \caption{The electromagnetic spectrum}    
        \endfirsthead
        \hline
        \multirow{2}{*}{$\displaystyle\lambda < \SI{0.1}{\angstrom}$}
         & Gamma radiation. \\
         &"You become Hulk" - Emiliano Descorvi, 2021.\\
        \hline    
        \multirow{3}{*}{$\displaystyle\lambda < \SI{10}{\angstrom}$}
         & X-rays. \\
         & Two types: soft and hard, hard being the more energetic. \\
         & Usually the result of the deexitisation of a radioactive nucleide. \\
        \hline
        \multirow{3}{*}{$\displaystyle\lambda < \SI{400}{\nano\metre}$}
         & Ultraviolet radiation. \\
         & Three types: A, B, and C, C being the more energetic. \\
         & Usually the result of the deexitisation of an electron.\\
        \hline
        \multirow{2}{*}{$\displaystyle\lambda < \SI{700}{\nano\metre}$}
         & Visible light. The wavelength limit is approximate, \\
         & as the visible spectrum varies for different people.\\
        \hline
        \multirow{3}{*}{$\displaystyle\lambda < \SI{1}{\milli\metre}$}
         & Infrared radiation. \\
         & Three types: near, mid, and far, near being the more energetic. \\
         & Usually caused by molecular vibrations.\\
        \hline
        \multirow{2}{*}{$\displaystyle\lambda < \SI{100}{\milli\metre}$}
         & Microwaves.\\
         & Not actually micro-scale.\\
        \hline
        \multirow{2}{*}{else}
         & Radiowaves. \\
         & These can have an arbitrarily long wavelength.\\
        \hline
    \end{longtable}

\subsection{Retarded potentials}
    Retarded potentials are the $\vec{A}$- and $V$-fields generated by a time-varying current (or charge distrubution) in the past, 
    since the fields propagate with the speed of light and not instantly. 
    
    By applying the \textit{Lorenz Gauge} $\nabla\cdot\vec{A} + \e\mu_0\frac{\partial V}{\partial t} = 0$, 
    Maxwell's equations for potentials become
    \begin{align}
        \dalambert V &= \frac{\rho\left(\vec{r}, t\right)}{\e} \\
        \dalambert \vec{A} &= \mu_0 \j\left(\vec{r}, t\right),
    \end{align}
    which are solved by 
    \begin{align} 
        \label{retarded potentials 1}
        V(\vec{r}, t) &= \frac{1}{4\pi\e} \iiint_\tau \frac{\rho\left(\vec{r'}, t_r\right)}{|\vec{r} - \vec{r'}|}d\tau \\
        \vec{A}(\vec{r}, t) &= \frac{\mu_0}{4\pi} \iiint_\tau \frac{\j\left(\vec{r'}, t_r\right)}{|\vec{r} - \vec{r'}|}d\tau,
        \label{retarded potentials 2}
    \end{align}
    where $\tau$ is the volume where $\rho$ and \j{} are defined, 
    and we integrate over all points $\vec{r'}$ in $\tau$.
    $t_r = t - \frac{|\vec{r} - \vec{r'}|}{c}$ is the \textit{retarded time}. 
    The notion of retarded time is to take the propagation speed into account.

\subsection{Dipoles as electromagnetic wave sources}
    In this chapter, an example of retarded potentials is discussed. 

    \subsubsection*{Introduction}
        If a charge $-q_0$ is placed at the origin of a refrence frame, 
        and a charge $q$ moves such that $\vec{r}(t) = z_0 \sin\left(\omega t\right) \hat{z}$, 
        then this is an oscilating dipole with dipole momentum 
        $\vec{p}\left(t\right) = q_0 z_0 \sin\left(\omega t\right)\hat{z} = p_0 \sin\left(\omega t\right)\hat{z}$.

        This configutarion is implactical, but a conductor with an oscilating current behaves similarly, since
        \begin{equation*}
            I(t) = \frac{\partial q}{\partial t} = \frac{1}{z_0} \frac{\partial p}{\partial t} = \frac{p_0}{z_0}\omega\cos\left(\omega t\right).
        \end{equation*}
    
    \subsubsection*{Finding the retarded potentials}
        Dimensional analysis of equation \eqref{retarded potentials 2}, along with the simplification $\vec{r} - \vec{r}' \approx \vec{r}$, 
        somehow leads to 
        \begin{equation}
            \vec{A}\left(\vec{r}, t\right) = \frac{\mu_0}{4\pi r} \frac{\partial }{\partial t}\vec{p}\left(t'\right),
        \end{equation} 
        where $t' = t - \frac{{r}}{c}$ is a simplified retarded time ($r = |\vec{r}|$).

        The potential is found by applying the Lorenz Gauge again:
        \begin{equation}
            V(\vec{r}, t) = \frac{1}{4\pi\e}
            \left[\frac{1}{cr}\frac{\partial}{\partial t} + \frac{1}{r^2} \right]
            p\left(t'\right),
        \end{equation}
        where $p = |\vec{p}|$ and $r = |\vec{r}|$. The expression in brackets is an operator, working on $p$.

    \subsubsection*{Arriving at the electromagnetic fields}
        We can now find \B{} and \E, since $\B = \nabla\times\vec{A}$ and $\E = -\nabla V - \frac{\partial\vec{A}}{\partial t}$:
        \begin{align*}
            E_r &= \frac{2\cos\theta}{4\pi\e r}\left[\frac{1}{cr}\frac{\partial}{\partial t} + \frac{1}{r^2}\right]
                    \eval{p\left(t\right)}_{t=t'}\\
            E_\phi &= 0 \\
            E_\theta &= \frac{\sin\theta}{4\pi\e r}\left[\e\mu_0\frac{\partial^2}{\partial t^2} 
                    + \frac{1}{cr}\frac{\partial}{\partial t} + \frac{1}{r^2}\right]
                    \eval{p\left(t\right)}_{t=t'}\\
            &\\
            B_r &= 0 \\
            B_\phi &= \frac{\mu_0 \sin\theta}{4\pi r}\left[\frac{1}{c}\frac{\partial^2}{\partial t^2} + \frac{1}{r}\frac{\partial}{\partial t}\right]
                \eval{p\left(t\right)}_{t=t'}\\
            B_\theta &= 0,
        \end{align*}
        where we still have $p = |\vec{p}| = p_0\sin\left(\omega t\right)$, $r = |\vec{r}|$, and $t' = t - \frac{{r}}{c}$. 
        The expressions in brackets are operators, working on $p(t)$. Note that $\theta$ is the azimuthal angle, and $\phi$ is the polar angle
        ($\theta \in [0, \pi]$, $\phi \in [0, 2\pi)$).

        For large $r$, the $r^{-2}$-terms vanish, leaving us with 
        \begin{align*}
            E_r &= 0\\
            E_\phi &= 0 \\
            E_\theta &= \frac{\sin\theta}{4\pi\e r c^2}\eval{\frac{\partial^2}{\partial t^2} p\left(t\right)}_{t=t'}\\
            &\\
            B_r &= 0 \\
            B_\phi &= \frac{\mu_0 \sin\theta}{4\pi r c^2}\eval{\frac{\partial^2}{\partial t^2} p\left(t\right)}_{t=t'}\\
            B_\theta &= 0,
        \end{align*}
        i.e. we have a spherical wave propagating in the $\hat{r}$-direction. 

    \subsubsection*{Jefimenko's equations (NOT part of the curriculum)}
        As a sidenote, if we apply the relations $\B = \nabla\times\vec{A}$ and $\E = -\nabla V - \frac{\partial\vec{A}}{\partial t}$
        directly to the general solutions found in equations \eqref{retarded potentials 1} and \eqref{retarded potentials 2}, we arrive at 
        \textit{Jefimenko's equations}:
        \begin{align*}
            \E\left(\vec{r}, t\right) &= \frac{1}{4\pi\e}\iiint_\tau \left[
                \frac{\vec{r} - \vec{r'}}{|\vec{r} - \vec{r'}|^3}\rho\left(\vec{r'}, t_r\right)
                + \frac{\vec{r} - \vec{r'}}{c|\vec{r} - \vec{r'}|^2}\frac{\partial}{\partial t}\rho\left(\vec{r'}, t_r\right)
                - \frac{1}{c^2|\vec{r} - \vec{r'}|} \frac{\partial}{\partial t}\j\left(\vec{r'}, t_r\right)
                \right]d\tau \\
            \B\left(\vec{r}, t\right) &= -\frac{\mu_0}{4\pi} \iiint_\tau \left[
                \frac{\vec{r} - \vec{r'}}{|\vec{r} - \vec{r'}|^3} \times \j\left(\vec{r'}, t_r\right)
                + \frac{\vec{r} - \vec{r'}}{|\vec{r} - \vec{r'}|^2} \times \frac{1}{c}\frac{\partial}{\partial t}\j\left(\vec{r'}, t_r\right)
                \right]d\tau,
        \end{align*}
        where $\rho$ and \j{} are arbitrary charge- and current density distributions, 
        $t_r = t - \frac{|\vec{r} - \vec{r'}|}{c}$ is the retarded time,
        $\tau$ is the volume where $\rho$ and \j{} are defined, 
        and we integrate over all points $\vec{r'}$ in $\tau$. 
        Please excuse the width of the equation for \E{}.

        These relations describe the electromagnetic fields for any arbitrary charge- and current density distributions, 
        taking the non-instantaneous propagation into account,
        and as such are very powerful if we are able to evaluate the integrals.
    
    \subsubsection*{Pointing vector and power}
        From our previous simplification, we get
        \begin{equation}
            \vec{S} = \frac{1}{\mu_0}\E\times\B = \frac{p^2\omega^2\sin^2\theta}{16\pi^2\e r^2 c^3}\hat{r},
        \end{equation}
        where the time-dependence is hidden in $p = p_0\sin\left(\omega\left(t - \frac{r}{c}\right)\right)$.
    
        The intensity is given as 
        \begin{equation}
            I = \left<\vec{S}\right> = \frac{p_0^2\omega^2\sin^2\theta}{32\pi^2\e r^2 c^3},
        \end{equation}
        since $\left<p^2\right> = \frac{1}{2}p_0^2$, and as such the intensity only depends on the $\theta$-coordinate. 
        Therefore, the flux of power through a sphere of any radius around the source is equal, as expected from conservation laws. 

        Finding the power propagating through a sphere $S$ around the dipole, we need to perform the integral 
        \begin{equation*}
            \iint_S \frac{p_0^2\omega^2}{32\pi^2\e c^3} \frac{\sin^2\theta}{r^2} dS = \iint_S I_0\frac{\sin^2\theta}{r^2} dS,
        \end{equation*}
        where the substitution of $dS = r^2\sin\theta d\phi d\theta$ yields 
        \begin{equation}
            P = \frac{p_0^2\omega^2}{12\pi\e c^3} = \frac{8\pi}{3}I_0,
        \end{equation}
        where $p_0$ is the amplitude of the dipole moment, and $\omega$ is the angular frequency of the dipole's oscilations.

        A unit analysis leads us into the \textit{Larmor formula}:
        \begin{equation}
            P = \frac{q^2a^2}{6\pi\e c^3},
        \end{equation}
        where $a$ is the acceleration of a charge $q$. This formula is only valid for non-relativistic speeds. 
        The Larmor formula predicts a constant power loss of any accelerating charge, radiating the energy away as electromagnetic waves.

\subsection{Diffusion of radiation}

    For a diluted dielectric medium, e.g. a gas of dielectric particles, the dipole momentum of each particle is 
    \begin{equation}
        \vec{p} = \e\alpha\E,
    \end{equation}
    where $\alpha$ is the polarisability of the medium, discussed in chapter \ref{Polarisation vector}, 
    and \E{} is the field produced by an incoming electromagnetic wave. 

    Since \E{} varies with time, $\vec{p}$ will as well, and as such power will be dissapated: 
    \begin{equation}
        P = \frac{4}{3}\pi\e\frac{\alpha^2 E_0^2}{\lambda^4},
    \end{equation}
    where $\lambda$ is the wavelength of the \textit{incoming} electromagnetic radiation. 
    This means that lower wavelengths are more scattered than higher wavelengths. 
    A very visible example of this is the blue colour of the sky.

\subsection{Propagation of EM waves in a medium}
    Here, a simple model for matter-electromagnetic wave-interractions is discussed. 

    Assume a dielectric material consists of elementary constituents, 
    each behaving like a damped oscilator oscilating along the $z$-axis.
    We formulate Newton's second law of motion as a differential equation: 
    \begin{equation*}
        -\frac{qE}{m} = \frac{d^2z}{dt^2} + \gamma\frac{dz}{dt} + \omega_0^2z,
    \end{equation*}
    where $-q$ is the charge of the particle, $m$ is the mass, $E$ is the magnitude of the electric field, 
    $\gamma$ is a damping factor, taking all contributions proportional to the velocity into account, 
    and $\omega_0^2 = \frac{k}{m}$ is the eigenfrequency of the oscilator ($k$ is the spring coefficient).

    Using complex wave notation for the electric field of the electromagnetic wave, 
    i.e. $\E = \Re\left[\E_0 e^{i\omega t}\right]$, we can write the solution as 
    \begin{equation}
        z(t) = \Re\left[\frac{-qE_0}{m\left(\omega_0^2 - \omega^2-i\gamma\omega\right)}e^{i\omega t}\right].
    \end{equation}
    We can now multiply by $q$ to get the dipole momentum, and further insert that into equation \eqref{polarisability} to arrive at
    \begin{equation}
        \alpha(\omega) = \frac{q^2}{\e m\left(\omega_0^2 - \omega^2 - i\gamma\omega\right)},
    \end{equation}
    i.e. the polarisability is frequency-dependent. Note that the polarisability can be complex, due to the $i\gamma\omega$-term.

    We can now find the polarisation vector, by using equation \eqref{polarisability} again: 
    \begin{equation}
        \p = \Re\left[\frac{nq^2}{\e m\left(\omega_0^2 - \omega^2 - i\gamma\omega\right)}\E_0e^{i\omega t}\right].
    \end{equation}
    Since $\alpha$ can be complex, \p{} and \E{} might be out of phase.

    This result can be further combined with equation \eqref{P as surface density}, yielding
    \begin{equation}
        \epsilon_r(\omega) = 1 + n\alpha(\omega),
    \end{equation}
    i.e. the relative dielectric constant is frequency-dependent and complex. 

\subsection{Refractive index}

    In a dielectric, the electric field is different than in a vacuum, and Maxwell's equations have an $\epsilon_r$ accompanying each \e. 
    The wave solution for these equations predict a propagation velocity of $v^{-2} = \e\mu_0\epsilon_r$, which is lower than $c$. 
    
    We define the \textit{refractive index} $n$ as follows: 
    \begin{equation}
        n(\omega) = \frac{c}{v(\omega)} = \Re\left[\sqrt{\epsilon_r(\omega)}\right],
    \end{equation}
    where $c$ is the speed of light, $v$ is the phase velocity
    in a medium with relative dielectric coefficient $\epsilon_r$. Note that $n \geq 1$.

    A material with refractive index $n > 1$ will have different wave parameters than if the same wave was propagating in vacuum: 
    \begin{align*}
        v &= \frac{c}{n} \\
        \lambda &= \frac{\lambda_0}{n} \\
        \nu &= \frac{v}{\lambda} = \frac{c}{\lambda_0} = \nu \\
        k &= nk_0,
    \end{align*}
    where a subscript $0$ denotes the relevant property in vacuum. $\nu = \frac{\omega}{2\pi}$ is frequency, and unchanged.

    So far, we have only discussed $n=\Re\left[\sqrt{\epsilon_r}\right]$, instead of the more general $\underline{n} = \sqrt{\epsilon_r}$.
    For the rest of the discussion, splitting the complex refractive index $\underline{n}$ into its real and imaginary parts is useful: 
    \begin{align}
        n = \Re\underline{n} &= 1 + \frac{n'q^2}{2\e m} \frac{\omega_0^2 - \omega^2}{\left(\omega_0^2 - \omega^2\right)+\gamma^2m^2} \\
        \kappa = \Im\underline{n} &= \frac{n'q^2}{2\e m} \frac{\gamma\omega}{\left(\omega_0^2 - \omega^2\right)+\gamma^2m^2},
    \end{align}
    where $n'$ is the density of elementary constituents (usually just $n$, marked here to differentiate it from refractive index),
    $q$ is the charge of the dipole each elementary constituent consists of, 
    $m$ is the mass of the atom/ molecule that makes up the dipole, 
    $\omega_0$ is the eigenfrequency of each constituent (it is assumed all constituents are the same, and therefore have the same $\omega_0$),
    $\omega$ is the angular frequency of the wave, 
    and $\gamma$ is a dissipative term arising from our simple damped oscilations model.
    The usefulness of this notation is mostly to describe attenuation: 
    \begin{align*}
        \E(z, t) &= \Re\left[\E_0 e^{i\left(\underline{k}z - \omega t\right)}\right] \\
        &= \Re\left[\E_0 e^{i\left(\underline{n}k_0z - \omega t\right)}\right] \\
        &= e^{-\kappa kz}\cdot\Re\left[\E_0 e^{i\left(nk_0z - \omega t\right)}\right], \numberthis \\
    \end{align*}
    i.e. we get an exponential decay in amplitude over distance. 

    The example used in the lectures was not of the electric field, but the intensity, which is proportional to the square amplitude of the fields: 
    \begin{equation}
        I = I_0 e^{-\beta z},
    \end{equation}
    where $\beta$ is the \textit{absorption coefficient} defined as
    \begin{equation}
        \beta = \frac{2\omega\kappa}{c} = 2\kappa k.
    \end{equation}

\subsection{Refraction and reflection}

    A monochromatic plane wave (i.e. a wave with a single frequency, and not a wave packet) 
    encountering an interface between two materials with different (purely real) refractive indices, will, in general, both reflect and refract. 
    The reflected wave will have the same angle $\theta_r$ with the normal to the interface as the angle $\theta_i$ the incident wave had, 
    but the transmitted (refracted) wave will have a new angle $\theta_t$. 

    The angles belonging to the same dielectric are the same since the phase of the radiation, 
    $|\vec{k}|\cdot\sin\theta$, must be equal for both waves. 

    A similar argument holds for the transmitted wave, and properly incorporating the refractive indices into $\vec{k}$ leads us into
    \textit{Snell's law}:
    \begin{equation}
        n_i\sin\theta_i = n_t\sin\theta_t.
    \end{equation}

    In general, the normal component of $\vec{k}$ is conserved over interfaces.

    \subsubsection*{Total internal reflection}
        If the angle between a transmitted electromagnetic wave and the normal of a dielectric interface is greater than or equal to $\frac{\pi}{2}$, 
        then we get a phenomenon known as \textit{total internal reflection} (TIR for short). 
        Expressed algebraically, $\theta_t \geq \frac{\pi}{2}$.
        If this condition is fulfilled, no power is transmitted into the other medium, and all radiation stays in its original medium.

        The \textit{critical angle} is found by Snell's law:
        \begin{equation}
            \theta_c = \text{arcsin}\frac{n_2}{n_1},
        \end{equation}
        where $n_1$ is the refractive index of the material of the incident wave, and $n_2$ is the refractive index of the material of the transmitted wave.
        If the angle the incident wave makes with the normal of the interface is greater than the critical angle, then we get TIR. 
        Algebraically, we get TIR if $\theta_i \geq \theta_c$. 
